---
title: |
  | Exploratory Data Analysis
  | MODBUS / TCP
author: "Lisa MALIPHOL"
date: June 2015
output:
  pdf_document:
    keep_tex: yes
---
```{r, echo=FALSE, message=FALSE}
require(TRSbook)
require(data.table)
require(igraph)
require(ggplot2)
require(gplots)
require(lattice)
require(dplyr)
require(plyr)
require(reshape2)
require(RColorBrewer)
require(png)
require(grid)
require(gridExtra)
```

```{r, message=FALSE, warning=FALSE, echo=FALSE}
# Load data from save(sewTCPConvDT, sewEndPtsDT, file="sewConvEnd.Rda")
load("/home/lisa/rDia/sewConvEnd.Rda")

# Load data from save(sewModbusDT, mergedSewDT, file="sew.Rda")
load("/home/lisa/rDia/sew.Rda")

```

# Introduction

The following analysis was done using a pcap file created from a simulation using the new SCADA simulation box under normal conditions (no attacks):

capture_sew_20150617.pcap   
3.0 MB   
38,082 packets   
10 minutes   

# Analysis

## Protocols

```{r, echo=FALSE, message=FALSE}
protocol <- sewModbusDT$ip.proto
camembert(protocol)
rm(protocol)
invisible(dev.off())
```

Conversation End Points:

```{r, echo=FALSE, warning=FALSE, message=FALSE}
sewModbusDT[,.(count=.N), by=.(ip.src, ip.dst, mbtcp.modbus.unit_id)]

```

```{r, echo=FALSE, message=FALSE, warning, fig.align='center'}
g <- graph.data.frame(sewTCPConvDT[,.(Address.A, Address.B, Packets, Bytes, Duration)], directed=FALSE)
gAdjMtx <- get.adjacency(g)  # adjacency mtx
gAdj <- graph.adjacency(gAdjMtx, mode="undirected", weighted=TRUE)
V(gAdj)$size <- degree(gAdj)*40  # node size according to degree of centrality
eW <- E(gAdj)$weight  # edge weights
eW <- ceiling(eW/max(eW)*55)
eW[eW>10] <- 10
plot.igraph(gAdj, main="Graph of SCADA Network", edge.width=eW)
rm(g, gAdjMtx, gAdj, eW)
invisible(dev.off())
```

```{r, echo=FALSE, warning=FALSE, message=FALSE, fig.height=6, fig.align='center'}
ggplot(sewTCPConvDT, aes(x = factor(0), y = Bytes/1e6)) +
  geom_boxplot(fill="darkorchid3") + xlab("") + scale_x_discrete(breaks = NULL) +
  facet_grid(~Address.A) +
  ggtitle("Boxplot of Packet Size by Source") +
  ylab("MBytes") +
  theme( plot.margin = unit(c(4,2,1,2), "cm") )

invisible(dev.off())
```

## Packet Analysis

```{r, echo=FALSE, warning=FALSE, message=FALSE}

# Number of Packets by Source
bc1 <- ggplot(sewModbusDT, aes(x=ip.src)) + geom_bar(stat="bin", fill="#FF9999", colour="black")
bc1 <- bc1 + ggtitle("Bar Chart of Packets by Source") + ylab("Packet Count")
bc1 <- bc1 + theme(axis.text.x = element_text(angle=90),
                   plot.title = element_text(size=10))

# Number of Packets by Destination
bc2 <- ggplot(sewModbusDT, aes(x=ip.dst)) + geom_bar(stat="bin", fill="#56B4E9", colour="black")
bc2 <- bc2 + ggtitle("Bar Chart of Packets by Destination") + ylab("Packet Count")
bc2 <- bc2 + theme(axis.text.x = element_text(angle=90),
                   plot.title = element_text(size=10))

grid.arrange(bc1, bc2, ncol=2)
```

```{r, echo=FALSE, warning=FALSE}
rm(bc1, bc2)
```

\pagebreak

### MODBUS/TCP Request/Response Packet Statistics (Appendix 1)


MODBUS/TCP requests are identified by packets having destination port number 502
```{r, echo=FALSE}
requests<-sewModbusDT[tcp.dstport=="502"]
```

```{r, echo=FALSE, warning=FALSE, message=FALSE, fig.align='center'}
p1 <- ggplot(requests, aes(x=factor(0))) +
  geom_boxplot(aes(y=frame.time_relative), fill="aquamarine4") + xlab("") +
  scale_x_discrete(breaks = NULL)
    
p2 <- ggplot(requests, aes(x=factor(0))) +
  geom_boxplot(aes(y=frame.time_delta_displayed)) + xlab("") +
  scale_x_discrete(breaks = NULL)
    
p3 <- ggplot(requests, aes(x=factor(0))) +
  geom_boxplot(aes(y=frame.len)) + xlab("") + scale_x_discrete(breaks = NULL)
    
p4 <- ggplot(requests, aes(x=factor(0))) +
  geom_boxplot(aes(y=mbtcp.len)) + xlab("") + scale_x_discrete(breaks = NULL)
    
grid.arrange(p1, p2, p3, p4, ncol=2, nrow=2)

invisible(dev.off())
```

\pagebreak

MODBUS/TCP responses are identified by packets having source port number 502
```{r, echo=FALSE, warning=FALSE}
responses<-sewModbusDT[tcp.srcport=="502"]
```

```{r, echo=FALSE, warning=FALSE, message=FALSE, fig.align='center'}
p1 <- ggplot(responses, aes(x=factor(0))) +
  geom_boxplot(aes(y=frame.time_relative), fill="steelblue2") + xlab("") +
  scale_x_discrete(breaks = NULL)
    
p2 <- ggplot(responses, aes(x=factor(0))) +
  geom_boxplot(aes(y=frame.time_delta_displayed)) + xlab("") +
  scale_x_discrete(breaks = NULL)
    
p3 <- ggplot(responses, aes(x=factor(0))) +
  geom_boxplot(aes(y=frame.len)) + xlab("") + scale_x_discrete(breaks = NULL)
    
p4 <- ggplot(responses, aes(x=factor(0))) +
  geom_boxplot(aes(y=mbtcp.len)) + xlab("") + scale_x_discrete(breaks = NULL)
    
grid.arrange(p1, p2, p3, p4, ncol=2, nrow=2)

invisible(dev.off())
rm(p1,p2,p3,p4)
```

\pagebreak

```{r, echo=FALSE, warning=FALSE, message=FALSE}
ggplot(sewModbusDT, aes(x=frame.number)) +
  geom_line(aes(y=frame.time_relative, color="frame.time_relative"))+
  geom_point(aes(y=mbtcp.len, color="mbtcp.len")) +
  ggtitle("Scatterplot of Time Recorded \nMODBUS Data Length as a \nFunction of Frame Number") +
  theme(axis.text = element_text(size = 13),
#        plot.title = element_text(face="bold"),
        plot.margin = unit(c(1, 0, 1, 0), "cm")) +
  scale_colour_manual(name='', values=c('frame.time_relative'='olivedrab',
                                        'mbtcp.len'='firebrick'), guide='legend') +
  guides(colour = guide_legend(override.aes = list(linetype=c(1,0)
                                                   , shape=c(NA, 16))))
invisible(dev.off())
```

```{r, echo=FALSE, warning=FALSE, message=FALSE, fig.height=4.5, fig.width=4, fig.align='center'}
ggplot(sewModbusDT,  aes(x=factor(0), y=mbtcp.len)) + geom_violin() +
  xlab("") + scale_x_discrete(breaks = NULL) + coord_flip() +
  ggtitle("Boxplot of MODBUS/TCP Data Length") +
  theme(#plot.title = element_text(face="bold"),
        plot.margin = unit(c(0, 0, 2, 0), "cm"))

invisible(dev.off())
```

\pagebreak

## MODBUS/TCP Data[^1] Analysis 
[^1]: https://www.wireshark.org/docs/dfref/m/mbtcp.html

The following analysis was done over the previous dataset that has been processed
to merge response packet to the request packet of the same transaction. An additional
field "d" is the data field "resp.data" converted from a hex to a decimal value.
(Appendix 2)
\    
\    
\    


```{r, warning=FALSE, message=FALSE, echo=FALSE}
ggplot(mergedSewDT, aes(x=frame.time_relative, y=d, color=mbtcp.modbus.reference_num)) +
  geom_point() + facet_grid(mbtcp.modbus.reference_num~resp.func.code) +
  ggtitle("MODBUS Data Values \nOver Time by Reference Number") + 
  scale_fill_brewer(palette="Spectral") +
  theme(#plot.title = element_text(face="bold"),
        plot.margin = unit(c(0, 0, 2, 0), "cm"),
        legend.position = "top", legend.title = element_text(size=10))
invisible(dev.off())
```

```{r, echo=FALSE, message=FALSE, warning=FALSE}
ggplot(mergedSewDT, aes(resp.time.rel, d, color=mbtcp.modbus.reference_num)) +
  geom_point() + facet_grid(~resp.func.code) +
  scale_fill_brewer(palette="Set1") +
  ggtitle("MODBUS Data Value Over Time by Function Code") +
  theme(#plot.title = element_text(face="bold"),
        plot.margin = unit(c(0, 0, 2, 0), "cm"),
        legend.position = "top", legend.title = element_text(size=10))
invisible(dev.off())
```

\pagebreak

## MODBUS Data Value Statistics
```{r, echo=FALSE, warning=FALSE, message=FALSE}
mergedSewDT[,.(count=.N, d.min=min(d), d.mean=mean(d, na.rm=T), d.max=max(d),
            d.sd=sd(d, na.rm=T), min.resp.time.rel=min(resp.time.rel),
            min.resp.time.rel= max(resp.time.rel)),
            by =.(resp.func.code, mbtcp.modbus.reference_num)][
           order(resp.func.code, mbtcp.modbus.reference_num)]
```

```{r, echo=FALSE, warning=FALSE, message=FALSE}
# Frequency of Reference Numbers by Function Code
ggplot(sewModbusDT, aes(x=mbtcp.modbus.reference_num, fill=mbtcp.modbus.reference_num)) +
  geom_bar(stat="bin") +
  facet_grid(~mbtcp.modbus.func_code) +
  ggtitle("Bar Chart of Reference Numbers by Function Code") +
  scale_fill_brewer(palette="Set1")

invisible(dev.off())
```

\pagebreak

```{r, echo=FALSE, warning=FALSE, message=FALSE}
ggplot(mergedSewDT, aes(x = factor(0), y = d)) +
  geom_boxplot(fill="lightpink4") +
  facet_grid(mbtcp.modbus.func_code~mbtcp.modbus.reference_num) +
  xlab("") + scale_x_discrete(breaks = NULL) +
  ggtitle("Boxplot of Modbus Data Values by Function/Reference Number")

invisible(dev.off())
```

```{r, echo=FALSE, warning=FALSE, message=FALSE}
pairs(mergedSewDT[,.(mbtcp.len, mbtcp.modbus.func_code=as.numeric(mbtcp.modbus.func_code),
                     mbtcp.modbus.reference_num=as.numeric(mbtcp.modbus.reference_num) , d)])
```

```{r, echo=FALSE, message=FALSE, warning=FALSE}
cloud(resp.time.rel ~ d + mbtcp.modbus.reference_num | resp.func.code, data = mergedSewDT[resp.func.code=="4"],
      col.point = mergedSewDT$mbtcp.modbus.reference_num, pch= 19,
      xlab = "d", ylab = "refNum", zlab = "resp.time.rel",
      panel.aspect = 0.7,
      main="3D Scatterplot \nReference Number, Data Value Over Time for Function Code 4",
      key = list(points = list(pch = 19, col = seq_along(levels(mergedSewDT$mbtcp.modbus.reference_num))), 
                 text = list(levels(mergedSewDT$mbtcp.modbus.reference_num)), space = 'top',
                 columns = nlevels(mergedSewDT$mbtcp.modbus.reference_num)))
invisible(dev.off())
```


```{r, echo=FALSE, message=FALSE, warning=FALSE}

rm(cor.test.2.sample)
```

\pagebreak

# Appendix 1 - MODBUS/TCP Request/Response Packet Statistics

```{r, warning=FALSE}
summary(requests)
```

```{r, warning=FALSE}
summary(responses)
```

```{r, echo=FALSE}
rm(responses, requests)
```

\pagebreak

# Appendix 2 - Merged Request/Response Packet Statistics

```{r, warning=FALSE}
summary(mergedSewDT)
```
