---
title: |
  | Diateam: SCAD\@COPS
  | A Hybrid Network Intrusion Detection System
author: "Lisa MALIPHOL"
date: "July 2015"
output:
  pdf_document:
    toc: true
    fig_caption: true
    keep_tex: yes
  word_document: default
---
```{r, echo=FALSE, message=FALSE, warning=FALSE}
require(grid)
require(png)

# http://rmarkdown.rstudio.com/authoring_basics.html
# https://www.rstudio.com/wp-content/uploads/2015/03/rmarkdown-reference.pdf
```

# Introduction

## Project background

## Problem definition

## Paper organization

# SCADA Systems

## Terms

### ICS

### SCADA

### PLC

### RTU

### HMI

## Traffic characterization  

# Protocols

## TCP

## MODBUS/TCP  

# Common Attacks on SCADA

## Command Injection

## Response Injection

## Denial of Service  

# Intrusion Detection Systems

## Host IDS

## Network IDS

### Signature-based

### Anomaly-based  

# Techniques of Network Intrusion Detection

## Statistical

## Machine Learning

## Data Mining  

# Tools

## Wireshark[^1] - Network Traffic Analysis Tool
[^1]: https://www.wireshark.org/docs/wsug_html_chunked

Developed in 1997 by Gerald Combs originally named Ethereal, Wireshark is now an Open Source GNU project. It is a network packet analyzer, or “packet sniffer”, that captures and displays network packets.

Captured network packets are saved in the pcap file format and can be dissected and parsed by Wireshark in order to analyze its contents.  An important aspect of Wireshark is that of its passive/monitoring nature and so does not send, manipulate, or modify the data passing over the network.

An initial packet capture file was created over simulated network traffic using Wireshark. Using its export facilities, various files were created for further analysis, with information such as TCP endpoints, conversations, etc.

## TShark[^2]
[^2]: https://www.wireshark.org/docs/man-pages/tshark.html

Another tool from the Wireshark suite is the command-line tool similar to tcpdump is tshark, a network protocol analyzer. In addition to capturing packet data over a live network, it is also capable of analyzing packets from an existing capture file. TShark was used to parse out various pertinent variables pertaining to the Modbus/TCP application protocol enclosed in the packet data.


## UNIX Utilities
In order to further parse and transform the data, the UNIX utility tool sed, which supports the use of regular expressions, was also used.

## R - Statistical Tool[^3]
[^3]: http://www.r-project.org/

R is an Open Source programming language and environment used for statistical computing and graphics. Initially developed by John Chambers at Bell Labs as the S language in 1993, R was created as a freely available version under the GNU project by  Ross Ihaka and Robert Gentleman at the University of Auckland, New Zealand.

Maintained by the R Development Core Team and with an active and growing community, it provides various statistical and graphical creation capabilities available under most operating systems, and is extensible with numerous packages available.

## C++

## SQLite3

## MongoDB/MySQL  
    
# Data Source

# Exploratory Data Analysis 

# Statistical Measures/Features  

# Architecture  
  
## Process (Figure 1)

* Step 1: Data Acquisition During Normal Activity - From the IDS appliance, 
        sniff the network traffic, extract and store data in a database.
* Step 2: Statistical Analysis
    + 2.1 - Process data - Perform any transformation, filtering and data
            cleansing necessary.
    + 2.2 - Calculate and determine statistical measures.
    + 2.3 - Configure appliance with statistical parameters.
* Step 3: Detection Mode - Appliance is set to detection mode.
        
## Technical Architecture (Figure 2)
  
  Figure 1 - Process  
```{r, echo=FALSE}
setwd("~/rDia")
img <- readPNG("imgs/process.png")
grid.raster(img)
```

  Figure 2 - Technical Architecture  
```{r, echo=FALSE}
setwd("~/rDia")
img <- readPNG("imgs/techArchitecture.png")
grid.raster(img)
```


# Implementation  
  

# Testing/Evaluation  
  

# Conclusion/Future Work  

  
# Glossary


# Bibliography
  

# Appendices


