---
title: |
  | Diateam: SCAD\@COPS
  | A Hybrid Network Intrusion Detection System
author: "Lisa MALIPHOL"
date: July 2015
output:
  pdf_document:
    keep_tex: yes
---

```{r, echo=FALSE, message=FALSE, warning=FALSE}
require(grid)
require(png)
```
\  
\  

  * Introduction
    - Project background
    - Problem definition
    - Paper organization
    
  * SCADA Systems
    - Terms
        * ICS
        * SCADA
        * PLC
        * RTU
        * HMI
    - Traffic characterization  
    
  * Protocols
    - TCP
    - MODBUS/TCP  
    
  * Common Attacks on SCADA
    - Command Injection
    - Response Injection
    - Denial of Service  
    
  * Intrusion Detection Systems
    - Host IDS
    - Network IDS
        * Signature-based
        * Anomaly-based  
        
  * Techniques of Network Intrusion Detection
    - Statistical
    - Machine Learning
    - Data Mining  
    
  * Tools
    - Wireshark
    - TShark
    - UNIX unitilites - sed, awk, bash, etc.
    - R
    - C++
    - SQLite3/MySQL  
    
  * Data Source
    
  * Exploratory Data Analysis 
  
  * Statistical Measures/Features  
  
  * Architecture  
  
    - Process (Figure 1)
        + Step 1: Data Acquisition During Normal Activity - From the IDS appliance, 
        sniff the network traffic, extract and store data in a database.
        + Step 2: Statistical Analysis
            * 2.1 - Process data - Perform any transformation, filtering and data
            cleansing necessary.
            * 2.2 - Calculate and determine statistical measures.
            * 2.3 - Configure appliance with statistical parameters.
        + Step 3: Detection Mode - Appliance is set to detection mode.
        
      - Technical Architecture (Figure 2)
  
  Figure 1 - Process  
```{r, echo=FALSE}
setwd("~/rDia")
img <- readPNG("imgs/process.png")
grid.raster(img)
```



  Figure 2 - Technical Architecture  
```{r, echo=FALSE}
setwd("~/rDia")
img <- readPNG("imgs/techArchitecture.png")
grid.raster(img)
```

  * Implementation  
  
  * Testing/Evaluation  
  
  * Conclusion/Future Work  
  
  * Bibliography
  
  * Appendices
  
